---
title: "ReprodResearch1"
author: "CheccaAird"
date: "13 April 2016"
output: html_document
---

Assignment 1 

Loading and preprocessing the data

```{r, echo=TRUE}
#Load the activity CSV data
activityData <- read.table("activity.csv", sep = ",", header = TRUE)
```


What is mean total number of steps taken per day?

```{r, echo=TRUE}
# Total number of steps per day 
stepsPerDay <- aggregate(steps ~ date, activityData, sum, na.rm= TRUE)

# Histogram of total number of steps taken per day 
hist(stepsPerDay$steps,breaks=5 , main = "Number of steps per day", col="green", xlab="Steps")

# Mean of steps, ignoring the missing values in the data set 
stepsPerDayMean <- mean(activityData$steps, na.rm= TRUE)
stepsPerDayMean

# Median of the total steps per day 
stepsPerDayMedian <- median(activityData$steps, na.rm= TRUE)
stepsPerDayMedian
```


What is the average daily activity pattern?

```{r, echo=TRUE}
# Time series plot of the 5 minute interval and the average number of steps taken, averaged across all days 
stepsdata <- aggregate(steps ~ interval, activityData, mean, na.rm=TRUE)
plot(stepsdata$interval, stepsdata$steps, type="l", main="Average Steps per Five Minute Interval", xlab="Interval No.", ylab="steps taken")
# The 5 minute interval with the maximum average number of steps 
maxsteps <- max(stepsdata$steps)
maxsteps

```


Within the data there a number of days and intervals that are missing values. This can introduce bias into some of the calculations and summaries, which is why I have devised a strategy for filling in the missing values in the dataset. 
```{r, echo=TRUE}
# The number of missing values in the dataset 
nullValues <- sum(is.na(activityData$steps))
nullValues
```
My strategy is to use the mean values from the dataset to fill in the missing values. While this is a basic method, it will serve for the purpose of this assignment as it should have the least impact on the data summaries. 

```{r, echo=TRUE}
# Fill in missing values 
activityData$steps[is.na(activityData$steps)] <- mean(activityData$steps, na.rm = TRUE)
# Dataset with the missing data filled in 
completeData <- aggregate(steps ~ date + interval, activityData, sum, na.rm= TRUE)
# Histogram of the total number of steps taken each day
hist(completeData$steps, breaks = 20, main = "Complete data- Total steps per day")
# Mean and median  of total number of steps taken per day 
completeDataMean <- mean(completeData$steps)
completeDataMean
completeDataMedian <- median(completeData$steps)
completeDataMedian
```
As I used the mean value for the missing values, there has been no impact on the median or the mean, which was my goal for filling in missing values. 


Are there differences in activity patterns between weekdays and weekends?

The lattice library is needed for this exercise. 

```{r, echo=TRUE}
library(lattice)
```


```{r, echo=TRUE}
# Factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day
completeData$date <- as.Date(completeData$date)
completeData$dayname <- weekdays(completeData$date)
completeData$weekend <- as.factor(ifelse(completeData$dayname == "Saturday" | completeData$dayname == "Sunday", "weekend", "weekday"))
# Panel plot contain 2 times series plots of the 5 minute interval and the average number of steps 
latticeData <- aggregate(steps ~interval + weekend, completeData, mean)
xyplot(steps ~ interval | factor(weekend), data=latticeData, type= "l", aspect = 1/3)
```



